---
title: "Loading, Merging and Joining Datasets"
subtitle: "Includes general housekeeping tasks like sorting variables names, creating essential features and sorting out variables order"
author: "Diego Usai"
date: "23 September 2019"
output:
  html_document:
    theme: spacelab
    df_print: paged
    highlight: pygments
    number_sections: false
    toc: true
    toc_float: true
    toc_depth : 4
    font-family: Roboto
    code_folding: none
    keep_md: false
    dpi: 300
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = TRUE,      # TRUE to evaluate every single chunck
  warning = FALSE,     # FALSE to suppress warnings from being shown
  message = FALSE,     # FALSE to avoid package loading messages
  cache   = FALSE,     # TRUE to save every single chunck to a folder
  echo    = TRUE,      # TRUE for display code in output document
  out.width = "80%",
  out.height = "80%",
  fig.align = "center"
)
```

```{r switch off locale, include=FALSE}
# turn off locale-specific sorting for messages in English
Sys.setlocale("LC_TIME", "C")
```

```{r Load libraries, message = FALSE, include=TRUE}
library(tidyverse)
library(lubridate)
library(readr)
```

## Outline

This is the minimal coding necessary to assemble the various data feeds and sort out the likes of variables naming & new features creation plus some general housekeeping tasks. I will continue to add to this code should the need arise for other features to be created.  


## The Dataset

The dataset I'm using here accompanies a Redbooks publication called [_Building 360-Degree Information Applications_](https://www.redbooks.ibm.com/abstracts/sg248133.html?Open) which is available as a free PDF download. Also available as a free download are the excel files in the [__Additional Material__](ftp://www.redbooks.ibm.com/redbooks/SG248133) section to follow along with the exercises in the book.


The data covers _3 & 1/2 years_ worth of sales `orders` for the __Sample Outdoors Company__, a fictitious B2B outdoor equipment retailer enterprise. The data comes with details about the `products` they sell as well as their customers (which are `retailers` in their case). The data is a subset of the _GSDB database_, a realistic and feature-rich database created by IBM® to support demos for their products.


## Load, merge and join

I will start with sales transactions as they are distributed across almost 900 separate excel files. 

First, I load and merge them at once from a single folder and bind them by row into an `orders` file .
```{r}
orders <- list.files(path = "../00_data/order-details",
           pattern = "*.csv", 
           full.names = T) %>% 
   map_df(~readr::read_csv(.))

```
__NOTE__ that all files have the same format with each single variable sitting on the same column in each file

```{r, collapse=TRUE}
orders %>% glimpse()

```


A bit of a clean up is required as there are 45 `orders` variables but most of them  are not necessary. For instance, `order_method` is available in many languages but I only need the one in English.

I start by removing all `order_method` other than English.
```{r}
orders <- 
   orders %>% 
   rename(
      order_method = order_method_en,
      retailer = retailer_name
      )  %>% 
   select(
      -contains('order_method_')
      ) 
```

```{r,, collapse=TRUE}
orders %>% glimpse()
```

Then I load the `products` file, which, as it says on the tin, contains the products information
```{r}
# load products file
products <- 
   read_csv('../00_data/products.csv')
```

```{r, collapse=TRUE}
products %>% glimpse()
```

Finally, the `retailers` file
```{r}
# load retailer info
retailers <- 
   read_csv('../00_data/retailers.csv')
```

```{r, collapse=TRUE}
retailers %>% glimpse()
```

As R is case sensitive, I'm setting `retailers` column names to lower-case
```{r}
retailers <- 
   retailers %>% 
   rename_all(tolower) 
```

Now I can joining `orders`, `products` and `retailers` info into one data frame
```{r}

df <- 
   orders %>% 
   left_join(products, by = ('product_number')) %>% 
   left_join(retailers, by = ('retailer_site_code')) 
```


## Creating essential features & some housekeeping tasks

One long piece of code to sort all in one go. 

I've added comments to each block to explain what it's doing. 
```{r}

orders_tbl   <- 
   
    # create revenue, total product cost and gross profit
   df %>%
      mutate(
         revenue       = quantity * unit_price,
         tot_prod_cost = quantity * unit_cost,
         gross_profit  = revenue - tot_prod_cost
      ) %>% 
      
      # replacing NAs in the return_count variable
   replace_na(list(return_count = 0)) %>% 
      
    # Shorten product line names for readablility and ease of use in charts and code
    
   mutate(
      prod_line = case_when(
         product_line == 'Camping Equipment' ~ 'Camping Eqpt',
         product_line == 'Golf Equipment' ~ 'Golf Eqpt',
         product_line == 'Mountaineering Equipment' ~ 'Mountain Eqpt',
         product_line == 'Personal Accessories' ~ 'Personal Acces',
         product_line == 'Outdoor Protection' ~ 'Outdoor Prot',
         TRUE ~ product_line
      ),
      
      prod_line_2 = case_when(
         product_line == 'Camping Equipment' ~ 'Camping Eqpt',
         product_line == 'Golf Equipment' ~ 'Golf Eqpt',
         product_line == 'Mountaineering Equipment' ~ 'Mountain Eqpt',
         product_line == 'Personal Accessories' ~ 'Personal Acces',
         product_line == 'Outdoor Protection' ~ 'Personal Acces',
         TRUE ~ product_line
      ),
      
      # create alternative regional group
      region2 = case_when(
         country_en %in% c('United Kingdom', 'France', 'Spain',  
                        'Netherlands','Belgium','Switzerland') ~ 'West Europe',
         country_en %in% c('Germany', 'Italy', 'Finland',  
                        'Austria','Sweden','Denmark') ~ 'East Europe',
         TRUE ~ region_en
      )
   ) %>% 
   
   # create financial years
   mutate(
      ord_date = ymd(order_date),
      fin_year = case_when(
         between(ord_date, ymd("2004-07-01"), ymd('2005-06-30')) ~ 'FY_04_05',
         between(ord_date, ymd("2005-07-01"), ymd('2006-06-30')) ~ 'FY_05_06',
         between(ord_date, ymd("2006-07-01"), ymd('2007-06-30')) ~ 'FY_06_07',
         TRUE ~ 'other'
      ),
      
      # create all quarters
      quarter_all = case_when(
         between(ord_date, ymd("2004-01-01"), ymd('2004-03-31')) ~ '04_Q1',
         between(ord_date, ymd("2004-04-01"), ymd('2004-06-30')) ~ '04_Q2',
         between(ord_date, ymd("2004-07-01"), ymd('2004-09-30')) ~ '04_Q3',
         between(ord_date, ymd("2004-10-01"), ymd('2004-12-31')) ~ '04_Q4',
         between(ord_date, ymd("2005-01-01"), ymd('2005-03-31')) ~ '05_Q1',
         between(ord_date, ymd("2005-04-01"), ymd('2005-06-30')) ~ '05_Q2',
         between(ord_date, ymd("2005-07-01"), ymd('2005-09-30')) ~ '05_Q3',
         between(ord_date, ymd("2005-10-01"), ymd('2005-12-31')) ~ '05_Q4',
         between(ord_date, ymd("2006-01-01"), ymd('2006-03-31')) ~ '06_Q1',
         between(ord_date, ymd("2006-04-01"), ymd('2006-06-30')) ~ '06_Q2',
         between(ord_date, ymd("2006-07-01"), ymd('2006-09-30')) ~ '06_Q3',
         between(ord_date, ymd("2006-10-01"), ymd('2006-12-31')) ~ '06_Q4',
         between(ord_date, ymd("2007-01-01"), ymd('2007-03-31')) ~ '07_Q1',
         between(ord_date, ymd("2007-04-01"), ymd('2007-06-30')) ~ '07_Q2',
         between(ord_date, ymd("2007-07-01"), ymd('2007-09-30')) ~ '07_Q3',
         TRUE ~ 'other'
      ),
      
      # create selected quarters
      quarter_sel = case_when(
         between(ord_date, ymd("2004-07-01"), ymd('2004-09-30')) ~ '04_Q3',
         between(ord_date, ymd("2004-10-01"), ymd('2004-12-31')) ~ '04_Q4',
         between(ord_date, ymd("2005-01-01"), ymd('2005-03-31')) ~ '05_Q1',
         between(ord_date, ymd("2005-04-01"), ymd('2005-06-30')) ~ '05_Q2',
         between(ord_date, ymd("2005-07-01"), ymd('2005-09-30')) ~ '05_Q3',
         between(ord_date, ymd("2005-10-01"), ymd('2005-12-31')) ~ '05_Q4',
         between(ord_date, ymd("2006-01-01"), ymd('2006-03-31')) ~ '06_Q1',
         between(ord_date, ymd("2006-04-01"), ymd('2006-06-30')) ~ '06_Q2',
         between(ord_date, ymd("2006-07-01"), ymd('2006-09-30')) ~ '06_Q3',
         between(ord_date, ymd("2006-10-01"), ymd('2006-12-31')) ~ '06_Q4',
         between(ord_date, ymd("2007-01-01"), ymd('2007-03-31')) ~ '07_Q1',
         between(ord_date, ymd("2007-04-01"), ymd('2007-06-30')) ~ '07_Q2',
         TRUE ~ 'other'
      )
      
      
   ) %>% 
   
   # reorder columns and rename a few
      select(
      order_number,
      order_date,
      order_close_date,
      order_ship_date = ship_date,
      fin_year,
      quarter_all,
      quarter_sel,
      order_method,
      retailer,
      retailer_code,
      retailer_type   = retailer_type_en,
      region          = region_en,
      region2,
      country         = country_en,
      city            = rtl_city,
      promotion_code,
      return          = return_count,
      quantity,
      unit_price,
      unit_sale_price,
      unit_cost,
      unit_prod_cost  = production_cost,
      unit_gross_marg = gross_margin,
      revenue,
      tot_prod_cost,
      gross_profit,
      prod_numb       = product_number,
      prod_line,
      prod_line_2,
      prod_type       = product_type,
      prod_name       = product_name,
      brand,
      color,
      size            = product_size,
      intro_date      = introduction_date,
      halt_date       = discontinued_date
   ) 
```

```{r, collapse=TRUE}
orders_tbl %>% glimpse()
```


## Remove original files and save

Last but not least, I can drop all original files and save the `orders_tbl`

```{r, evaluate=FALSE}
# remove original files
rm(df, products, retailers, orders)

# Save orders as RDS
orders_tbl %>%
   write_rds("../00_data/orders_tbl.rds")
```


### Code Repository
The full R code can be found on [my GitHub profile](https://github.com/DiegoUsaiUK/Loading_Merging_and_Joining_Datasets)

NOTE: given their size, the RDS file and the order-details folder had to be compress before uploading them on my Github profile





